Demystifying the Computer Science PhD Admission in the US\\{\large A Guide for International Students

### 2024-01-04 demystify.tex
updates

\begin{enumerate}
  \item Should you apply to a CS PhD in the US?
        \begin{itemize}
          \item \emph{Yes, definitely}.  CS PhD study in the US is fully funded and admission into good universities is not any harder than non-US schools (\S\ref{sec:should}).
        \end{itemize}
  \item How is your application evaluated?
        \begin{itemize}
          \item Applications are evaluated by the \emph{PhD Admission} committee and each application is reviewed by typically three faculty (\S\ref{sec:evalapps}).
          \item Individually faculty \emph{cannot directly admit} a student---so do not email and ask if you have a chance. However, faculty can \emph{advocate} for a student and therefore increase their admission chance---so contact and introduce yourself (\S\ref{sec:contact}).
        \end{itemize}
  \item Application Materials
        \begin{itemize}
          \item The committee will look at various factors, but the most important ones are research ability, e.g., publications, personal statement, and recommendation letters.
          \item LORs are very important, but only if they are personalized and talk about your research ability (\S\ref{sec:lor}).
          \item Personal statement is very important. Write it in such a way that makes you \emph{stand out} (\S\ref{sec:research-statement} and \S\ref{sec:improve-your-chance})
          \item GRE \emph{is not} required (\S\ref{sec:grades}). Spend your time on something else.
          \item Grades are important, but depend on the reputation of your school (\S\ref{sec:grades}).
          \item Getting an interview is typically a \emph{good sign} (no one interviews weak candidates).
        \end{itemize}
  \item What to do after getting admitted?
        \begin{itemize}
          \item Attend \emph{Open House} to learn more about the place and \emph{interview} profs---they would be much more willing to talk to you now (\S\ref{sec:accepted}).
        \end{itemize}
  \item Funding
        \begin{itemize}
          \item TA and RA are two main sources of funding.  TA (teaching assistantship) is provided by the department to help profs. with classes (e.g., grading). RA (research assistantship) is provided by profs. to help them with their research (\S\ref{sec:funding}).
        \end{itemize}
  \item Choosing School and Professors
        \begin{itemize}
          \item Many schools do not offer PhD studies in CS and many CS professors \emph{cannot} be adviser, i.e., cannot graduate PhD students  (\S\ref{sec:schoolsandprofs}).
          \item Contacting a prof. for research opportunities is recommended, but do it \emph{properly} (\S\ref{sec:contact}).
        \end{itemize}
  \item Miscs and FAQS
        \begin{itemize}
          \item Increasing your admission chance by being unique and standing out (\S\ref{sec:improve-your-chance}).
          \item You can successfully apply to CS PhD even if you have non-STEM background (\S\ref{sec:non-stem}).
          \item Compared to other countries, CS PhD in the US does not require an MS degree but has longer PhD study time (\S\ref{sec:non-us-differences}).
          \item Your PhD costs about \$400K in total, but you \emph{do not} pay for it (\S\ref{sec:ra-cost}).
          \item Despite some miserable stories on social media, many PhD students have good mentors, supportive lab mates, healthy working environment ... and are happy (\S\ref{sec:happy}).
        \end{itemize}
\end{enumerate}

@@

@@

\section{Academic Integrity}

@@

\chapter{Ranking of CS PhD programs}\label{sec:ranking}

\begin{table}
  \centering
  \small
  \caption{Top 50 CS PhD programs in the U.S. (CSRankings, June 2023). \red{$^*$} indicates that the university has \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Viet-CS-Profs-US}{Vietnamese prof.} that can advise CS PhD students.}\label{tab:ranking}
  \begin{tabular}{rl|rl}
    \toprule
    1 & Carnegie Mellon & 26 & Univ. of California - Irvine \\
    2 & Univ. of Illinois at Urbana-Champaign\red{$^*$}  & 27 &  Duke University \\
    3 & Univ. of California-San Diego & 28 & Rutgers University\red{$^*$} \\
    4 & Georgia Institute of Technology & 29 & Univ. of California - Riverside\\
    4 & MIT         & 30 & Northwestern University\\
    6 & Univ. of California - Berkeley& 31 & Pennsylvania State University  \\
    7 & University of Michigan - Ann Arbor\red{$^*$}   & 32& George Mason University\red{$^*$}\\  
    8 & University of Washington      &33 &  Harvard University \\
    9 &  Stanford University  &34&  Univ. of California - Santa Cruz \\
    10 & Cornell University  & 35 &  Yale University \\
    11 & University of Maryland - College Park &  36& Brown University \\ 
    12 & Northeastern University\red{$^*$} &37&  Ohio State University\\
    13 & University of Wisconsin - Madison\red{$^*$}  &38& Texas A\&M University\red{$^*$} \\
    13 & Purdue University  &39 & Boston University  \\
    15 & University of Texas at Austin &40& North Carolina State University\\\  
    16 & University of Pennsylvania &41 & University of Utah \\
    17 & Columbia University \red{$^*$} &42 & University at Buffalo\red{$^*$}\\
    18 & Princeton University  & 43& Rice University\\
    19 & New York University  & 44&  University of Colorado-Boulder \\
    20 &  University of Massachusetts-Amherst\red{$^*$} &45& University of Illinois at Chicago  \\
    21 & Univ. of California - Los Angeles &46& Virginia Tech\red{$^*$}  \\
    22 & University of Southern California &47&  Arizona State University\red{$^*$} \\
    23 & Stony Brook University\red{$^*$} &48&University of Minnesota \\
    24 & University of Chicago &49& University of Virginia \\
    25 & Univ. of California - Santa Barbara &50& University of North Carolina\red{$^*$} \\
    \bottomrule
  \end{tabular}
\end{table}

Tab.~\ref{tab:ranking} lists the top 50 CS programs in the US from \href{https://www.csrankings.org}{CSRankings.org}, a ranking system  based on top CS conferences.

@@

          %           \newpage
          %           \section{Virtual Research Opportunities Beyond Physical Boundaries (written by Didier)}
          %           \didi{In my experience, prospective PhD students (from Rwanda) are often unaware of the information provided in this document throughout their undergraduate. For this reason they often find themselves unprepared for PhD admission process by the time they finish undergraduate and have to spend a year or more to improve their application profiles. I think it would be beneficial to expand the document to also target students who are not yet ready for the application process to help them improve on all criteria considered during admission. I drafted the short essay on how to gain research experience remotely in order to improve their research profile and potentially get LOR from experts (hopefully faculty)}\\
          %           In the realm of computer science, research experience plays a crucial role in securing admissions to top-tier Ph.D. programs. However, students from underrepresented small colleges, both in the United States and internationally, often encounter limited research opportunities within their institutions. Thankfully, there exists an alternative pathway for these aspiring researchers to gain valuable research experience, even without being physically present at a university.

          %           \textbf{Online Research Communities and Open Source Contributions}:
          %           Online research communities and platforms offer a wealth of opportunities for aspiring researchers. Platforms like GitHub, GitLab, and Bitbucket host repositories spanning diverse research domains. Students can explore repositories related to their areas of interest, collaborate with other contributors, and actively engage in open source research projects. By contributing code, fixing bugs, implementing new features, or providing documentation, students can gain practical research experience and interact with experienced developers and researchers in the field.

>>>>

\begin{enumerate}
  \item Should you apply to a CS PhD in the US?
        \begin{itemize}
          \item \emph{Yes, definitely}.  CS PhD study in the US is fully funded and admission into good universities is not any harder than non-US schools (\S\ref{sec:should}).
        \end{itemize}
  \item How is your application evaluated?
        \begin{itemize}
          \item Applications are evaluated by the \emph{PhD Admission} committee and each application is reviewed by typically three faculty (\S\ref{sec:evalapps}).
          \item Individually faculty \emph{cannot directly admit} a student---so do not email and ask if you have a chance. However, faculty can \emph{advocate} for a student and therefore increase their admission chance---so contact and introduce yourself (\S\ref{sec:contact}).
        \end{itemize}
  \item Application Materials
        \begin{itemize}
          \item The committee will look at various factors, but the most important ones are research ability, e.g., publications, personal statement, and recommendation letters.
          \item LORs are very important, but only if they are personalized and talk about your research ability (\S\ref{sec:lor}).
          \item Personal statement is very important. Write it in such a way that makes you \emph{stand out} (\S\ref{sec:research-statement} and \S\ref{sec:improve-your-chance})
          \item GRE \emph{is not} required (\S\ref{sec:grades}). Spend your time on something else.
          \item Grades are important, but depend on the reputation of your school (\S\ref{sec:grades}).
          \item Getting an interview is typically a \emph{good sign} (no one interviews weak candidates).
        \end{itemize}
  \item What to do after getting admitted?
        \begin{itemize}
          \item Attend \emph{Open House} to learn more about the place and \emph{interview} profs---they would be much more willing to talk to you now (\S\ref{sec:accepted}).
        \end{itemize}
  \item Funding
        \begin{itemize}
          \item TA and RA are two main sources of funding.  TA (teaching assistantship) is provided by the department to help profs. with classes (e.g., grading). RA (research assistantship) is provided by profs. to help them with their research (\S\ref{sec:funding}).
        \end{itemize}
  \item Choosing School and Professors
        \begin{itemize}
          \item Many schools do not offer PhD studies in CS and many CS professors \emph{cannot} advise or graduate PhD students  (\S\ref{sec:schoolsandprofs}).
          \item Contacting a prof. for research opportunities is recommended, but do it \emph{properly} (\S\ref{sec:contact}).
        \end{itemize}
  \item Miscs and FAQS
        \begin{itemize}
          \item Increasing your admission chance by being unique and standing out (\S\ref{sec:improve-your-chance}).
          \item You can successfully apply to CS PhD even if you have non-STEM background (\S\ref{sec:non-stem}).
          \item Compared to other countries, CS PhD in the US does not require an MS degree but has longer PhD study time (\S\ref{sec:non-us-differences}).
          \item Your PhD costs about \$400K in total, but you \emph{do not} pay for it (\S\ref{sec:ra-cost}).
          \item Despite some miserable stories on social media, many PhD students have good mentors, supportive lab mates, healthy working environment ... and are happy (\S\ref{sec:happy}).
        \end{itemize}
\end{enumerate}

@@

\chapter{Rankings of CS PhD programs}\label{sec:ranking}

\begin{table}
  \centering
  \small
  \caption{Top 50 CS PhD programs in the U.S. (\href{https://csrankings.org}{CSRankings}, Jan. 2024). \red{$^*$} indicates that the university has \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Viet-CS-Profs-US}{Vietnamese prof.} that can advise CS PhD students.}\label{tab:ranking}
  \begin{tabular}{rl|rl}
    \toprule
    1 & Carnegie Mellon & 26 & Duke University \\
    2 & Univ. of Illinois at Urbana-Champaign\red{$^*$}  & 27 & Univ. of California - Santa Barbara \\  
    3 & Univ. of California-San Diego & 28 & Rutgers University\red{$^*$} \\
    4 & Georgia Institute of Technology & 29 & Univ. of California - Riverside\\
    5 & MIT                            & 30 & Pennsylvania State University  \\
    6 & Univ. of California - Berkeley& 31 & Northwestern University\\
    7 & University of Michigan - Ann Arbor\red{$^*$}   & 32& George Mason University\red{$^*$}\\  
    8 & University of Washington      &33 &  Texas A\&M University\red{$^*$} \\
    9 &  Stanford University  &34&  University of Utah \\
    10 & Cornell University  & 35 &  North Carolina State University\\\  
    11 & University of Maryland - College Park &  36& Ohio State \\ 
    12 & Northeastern University\red{$^*$} &37& University of Virginia  \\ 
    13 & Purdue University &38& Yale University \\ 
    14 & University of Wisconsin - Madison\red{$^*$} &39 & Univ. of California - Santa Cruz \\
    15 & University of Texas at Austin &40& Brown University \\ 
    16 & University of Pennsylvania &41 & Harvard University \\
    17 & Columbia University \red{$^*$} &42 & Boston University  \\
    18 & Princeton University  & 43& Rice University\\
    19 & New York University  & 44&  University at Buffalo\red{$^*$}\\
    20 &  University of Massachusetts-Amherst\red{$^*$} &45& University of Colorado-Boulder \\
    21 & Univ. of California - Los Angeles &46& University of Illinois at Chicago  \\
    22 & University of Southern California &47& University of Minnesota \\ 
    23 & University of Chicago &48& University of North Carolina\red{$^*$} \\
    24 & Stony Brook University\red{$^*$} &49& Arizona State University\red{$^*$} \\
    25 &  Univ. of California - Irvine&50& Univ. Of California - Davis \\
    \bottomrule
  \end{tabular}
\end{table}
%Virginia Tech\red{$^*$}  \\
Tab.~\ref{tab:ranking} lists the top 50 CS programs in the US from \href{https://www.csrankings.org}{CSRankings.org}, a ranking system  based on faculty publications at top CS conferences.

% \chapter{Undergraduate Research Opportunities}

% Prospective PhD students are often unaware of the importance of research experience in their application.  This provides both concrete research experiences and potential LOR writers that can talk about your research ability.
% %This is especially true for students from small colleges where research opportunities are limited.  
% This section aims to provide some guidance on how to gain research experience as an undergraduate student. %This section is written together Didier, a PhD student at GMU who did his undergraduate study in Rwanda.

% \paragraph{Locally} There are many ways to find research opportunities right in your own institution.  If you did well or like a class, you can check with the professor of that class for research opportunities.  Sometimes professors would mention that they look for undergraduates on their websites, so do check those out! Even if they say they are looking for graduate students, you still should contact them and ask.  You can also ask your academic adviser or other faculty members for suggestions.  Finally, you can also ask your peers who are already doing research.  You can use the method described in \S\ref{sec:contact} to contact professors

% In short, the most important thing is that you need to \emph{ask}!  Don't be shy, don't be afraid of rejection, and don't be afraid of being ignored or that you're not good enough.  Just ask!  The worst thing that can happen is that you get a ``no'' or no response.  But you will never know if you don't ask.  And if you get a ``no'', you can always ask again later.  There are students who got rejected the first time but got accepted the second time.  So don't give up!

% \paragraph{Online Research Communities and Open Source Contributions}:
%                     Places such as Github offer many opportunities for research, even though they might not appear so (e.g., they do not explicitly mention research). Students can explore repositories related to their areas of interest, collaborate with other contributors, and actively engage in open source research projects. By contributing code, fixing bugs, implementing new features, or providing documentation, students can gain practical research experience and interact with experienced developers and researchers in the field.

% \paragraph{Paid or Unpaid}  travel opportunities, publishing opportunities, etc

%                    \section{Virtual Research Opportunities Beyond Physical Boundaries (written by Didier)}
%                     \didi{In my experience, prospective PhD students (from Rwanda) are often unaware of the information provided in this document throughout their undergraduate. For this reason they often find themselves unprepared for PhD admission process by the time they finish undergraduate and have to spend a year or more to improve their application profiles. I think it would be beneficial to expand the document to also target students who are not yet ready for the application process to help them improve on all criteria considered during admission. I drafted the short essay on how to gain research experience remotely in order to improve their research profile and potentially get LOR from experts (hopefully faculty)}\\
%                     In the realm of computer science, research experience plays a crucial role in securing admissions to top-tier Ph.D. programs. However, students from underrepresented small colleges, both in the United States and internationally, often encounter limited research opportunities within their institutions. Thankfully, there exists an alternative pathway for these aspiring researchers to gain valuable research experience, even without being physically present at a university.

@@

\section{Accepting, Postponing, and Rejecting Offers}

Students, especially international ones, often get confused on what to do after they get an offer from a professor, e.g., how to postpone or accept/reject offers, and most importantly, how to do so without offending anyone. In short, if you are not serious about an offer for any reason (e.g., still waiting for other offers), do not accept it. If you are not sure, ask for more time to decide.  But do not accept an offer and then later reject it.  This is considered rude and unprofessional.  If you do so, you will likely burn bridges with that professor.

\paragraph{Accepting an offer} If you decide to accept an offer, you can say something like: \emph{``Thank you for the offer.  I am very excited about it and would like to accept it.  I look forward to working with you.  Can you please send me more details about the offer and what to do next?''} .  Here you are not only accepting the offer but also asking for more details.  This is perfectly fine and professors will understand and appreciate your enthusiasm.  They will likely send you more details about the offer and what to do next.  If you decide to accept an offer, you should do so quickly.

 If you have better offers, you might mention them to the professor and ask if they could match them. It is unlikely they can do so as RA might be fixed by the department and not negotiable. They might recommend you for departmental fellowships or some additional perks.

\paragraph{Postponing an offer} If you need more time to decide, you can ask for more time.  You can say something like: \emph{``Thank you for the offer.  I am very excited about it.  However, I am still waiting for other offers and need more time to decide.  Would it be possible to postpone the decision for a few weeks?''}.  This is perfectly fine and professors will understand and might even appreciate your honesty.  They will likely give you a few weeks to decide.  If you need more time, you can ask for more time.  But do not ask for too much time, e.g., more than a month.  You also should not postpone the offers multiple times, which will likely annoy people.

\paragraph{Rejecting an offer} If you decide to reject an offer, you can say something like: \emph{``Thank you for the offer.  I am very excited about it.  However, I have decided to accept another offer.  I appreciate your time and consideration.  I hope we can work together in the future.''}  Professors will understand and will likely wish you good luck.  If you decide to reject an offer, you should do so quickly.

This is very similar to a professional job offer.  You should be serious about it and communicate promptly. And most importantly, do not accept an offer and then later reject it.  If you do this with a company, you might face legal consequences.  If you do this with a professor, they will likely not do anything about it (i.e., not worth their time) but they will not want to work with you in the future.

%1. I'm fine if I extend an offer and you request a delay, even if it's because you're exploring other programs. No issue here, it's similar to job offerings. Of course, I may not be able to wait indefinitely and may extend the offer to another student, but I'll let you know that.
%2. I'm also okay if you find a program or professor that's a better fit. Similar to the first point, no hard feelings at all. However, I'd appreciate a prompt notification of your decision so I can extend an offer to another student.
%3. I'm NOT okay if you accept my offer but then, upon finding something better, tell me of your decision to withdraw Just like in job offers, once you are committed to an offer, there should be a compelling reason to withdraw (and claiming "I found somewhere better" is the weakest excuse).
%While I won't take any actions against you (a company might), I naturally will remember this and may be cautious in future interactions. I'll also reconsider making similar offers to students in similar situations.

\section{Academic Integrity (Cheating and Plagiarism)}

@@

@@
